Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)

### 2024-02-29 doc/manuscript/z_archive/ESSD/ForC-EFDB-manuscript-for-ESSD.tex
archive files

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned net emissions reductions
under the 2015 Paris Agreement. However, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accessible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
inventories if made available through the Emission Factor Database
(EFDB) of the International Panel on Climate Change (IPCC). Here, we
develop and document a process for semi-automated submission of data
from ForC into the EFDB, assess the data available and submitted to
date, and provide recommendations for improving forest data collection,
analysis, and reporting to improve inventories of forest-sector
greenhouse gas emissions and removals. We begin by reconciling
terminology and mapping ForC fields into EFDB. This process required
some updates to the ForC database structure, leading to the release of a
new version of ForC (v4.0; described here). As of June 09, 2023, ForC
contained \textasciitilde19316 independent records relevant to EFDB,
1438 of which have undergone necessary review and been submitted to
date. Among the data in ForC, there is disproportionate representation
of biomass (particularly aboveground) stocks, with far fewer records for
dead organic matter and soil C, and relatively few or no records for net
annual increments or C fluxes into or out of the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest
densities of relevant records in temperate forests, and with relatively
scant representation of tropical forests in Africa and Asia. In the
future, forest C estimates in EFDB can be improved through targeted
research to fill critical gaps, reporting of information required by
IPCC, and continued submission of data from scientific publications to
the EFDB. Given that climate change is rapidly impacting the world's
forests, timely reporting of recent estimates will be especially
critical to accurate forest C inventories.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, and the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
As of June 09, 2023, ForC contained 39848 records from 10589 plots in
1535 distinct geographical areas, along with records of stand age and
disturbance history. As such, ForC is positioned to improve
forest-related estimates of CO\textsubscript{2} emissions and removals
through the submission of data to the EFDB. The purpose of this
publication is to document that process and provide recommendations for
future improvements.

@@

To create structure for EFDB-relevant records, we added a total of 15
new variables to the set of named and defined variables (Fig. 1),
counting each pair of variables with units in C (ending in \texttt{\_C})
or organic matter (ending in \texttt{\_OM}) as one. The majority of
these were increment variables (n=11), adding to only one previously
defined increment variable (aboveground biomass increment,
\emph{delta.agb}). These are directly related to C stocks as previously
defined in ForC, with ``\emph{delta.}'' added in front of the variable
name. Further, we added variables capturing the belowground component of
woody mortality (\emph{woody.mortality\_root}) and the combined
aboveground and belowground components of woody mortality
(\emph{woody.mortality}). Although most of these variables lacked
records in ForC as of June 09, 2023, their addition gave the structure
such that records can be populated over time. Finally, to provide better
definition of the previously existing variable \emph{organic.layer},
which has a nebulous definition that reflects the varied definitions
adopted by original studies, we added two clearly defined variables:
\emph{litter} (relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

To submit complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data submission were that (1)
records had been checked against the original study and determined to be
complete and correct, and as originally presented, (2) the original
study presented values in tables or text, as opposed to the values
having been digitized from graphs or calculated based on related
variables, and (3) the records had not previously been submitted to
EFDB. Once converted into EFDB format, the records were reviewed and
then sent to the IPCC's Technical Support Unit for submission to EFDB.
Complete records needed to be reviewed by the EFDB editorial board and
then posted in the database -- a process that lags behind submission of
records and had not yet been completed for all records sent as of June
09, 2023.

@@

As of June 09, 2023, ForC (v4.0) contained 32686 independent records
(39848 total), 19316 of which were for the 42 variables relevant to EFDB
(Fig. 1). These records were distributed across all forested continents
and ecozones, albeit unevenly (Fig. 2). The largest number of records
came from Asia, followed by North America, South America, and Europe,
with relatively few records from Africa, Australia, and Oceania (Fig.
3c). Categorized by FAO ecozone, the greatest numbers of records came
from subtropical humid forests, temperate mountain systems, and tropical
rain forests, each with \textgreater2,000 independent records (Fig. 3b).
Boreal coniferous forests, temperate continental forests, subtropical
mountain systems, and tropical moist deciduous forests had
\textgreater1,000 independent records each, while other ecozones all had
\textless1,000 records. The most widely represented forest type was
needleleaf evergreen, followed by broadleaf deciduous and broadleaf
evergreen (Fig. 3a). In terms of stand age, the most represented age
class was 20-100 years, followed by \textless20 years and then
\textgreater100 years (Fig. 3d).

@@

As of June 09, 2023, we had reviewed or added 2292 EFDB-relevant
records, 1438 records of which were submitted to EFDB, and 376 of which
have been reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
attenuation between records reviewed and those sent to EFDB was
attributable to the presence of digitized records and records where a
variable's value had been calculated as the sum or difference of related
variables rather than presented directly in the text. The discrepancy
between the number of records sent and that posted to EFDB is primarily
attributable to the time required for the IPCC to review and post the
records, and also because a minority of records were deemed not
applicable to EFDB by the review panel.

@@

The ForC database contains large numbers of records that could
potentially be useful for estimating C stock changes applying
methodological guidance provided by the IPCC. Here we have developed a
framework for submitting these records to the EFDB, thus making those
data more accessible for reporting CO\textsubscript{2} emissions and
removals from forest land consistent with good practice in the IPCC
guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of June 09, 2023,
we have submitted 1438 records to EFDB. Although this represents just
7\% of relevant records in ForC, it substantially increases the number
of forest land records in EFDB. The records submitted to EFDB and
present in ForC are very unevenly distributed across variables, regions,
and forest types (Figs. 2-3, Table 2), reflecting broader patterns in
allocation of research effort.

@@

\codedataavailability{All code and data are openly available. The ForC
database and associated code are available via the ForC repository
within the ForC-db organization on GitHub
(https://github.com/forc-db/ForC), and the version used here (ForC v4.0)
is archived in Zenodo (DOI: TBD). The data and code associated with data
submission to EFDB and preparation of this manuscript are available via
the the IPCC-EFDB-integration repository within the ForC-db organization
on GitHub (https://github.com/forc-db/IPCC-EFDB-integration) and
archived in Zenodo (DOI:
10.5281/zenodo.8021474).} %% use this section when having data sets and software code available

>>>>

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned net emissions reductions
under the 2015 Paris Agreement. However, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accessible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
inventories if made available through the Emission Factor Database
(EFDB) of the International Panel on Climate Change (IPCC). Here, we
develop and document a process for semi-automated submission of data
from ForC into the EFDB, assess the data available and submitted to
date, and provide recommendations for improving forest data collection,
analysis, and reporting to improve inventories of forest-sector
greenhouse gas emissions and removals. We begin by reconciling
terminology and mapping ForC fields into EFDB. This process required
some updates to the ForC database structure, leading to the release of a
new version of ForC (v4.0; described here). As of July 28, 2023, ForC
contained \textasciitilde19316 independent records relevant to EFDB,
1438 of which have undergone necessary review and been submitted to
date. Among the data in ForC, there is disproportionate representation
of biomass (particularly aboveground) stocks, with far fewer records for
dead organic matter and soil C, and relatively few or no records for net
annual increments or C fluxes into or out of the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest
densities of relevant records in temperate forests, and with relatively
scant representation of tropical forests in Africa and Asia. In the
future, forest C estimates in EFDB can be improved through targeted
research to fill critical gaps, reporting of information required by
IPCC, and continued submission of data from scientific publications to
the EFDB. Given that climate change is rapidly impacting the world's
forests, timely reporting of recent estimates will be especially
critical to accurate forest C inventories.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021, anderson-teixeira_forcdb_2023}.
ForC includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, and the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
As of July 28, 2023, ForC contained 39848 records from 10589 plots in
1535 distinct geographical areas, along with records of stand age and
disturbance history. As such, ForC is positioned to improve
forest-related estimates of CO\textsubscript{2} emissions and removals
through the submission of data to the EFDB. The purpose of this
publication is to document that process and provide recommendations for
future improvements.

@@

To create structure for EFDB-relevant records, we added a total of 15
new variables to the set of named and defined variables (Fig. 1),
counting each pair of variables with units in C (ending in \texttt{\_C})
or organic matter (ending in \texttt{\_OM}) as one. The majority of
these were increment variables (n=11), adding to only one previously
defined increment variable (aboveground biomass increment,
\emph{delta.agb}). These are directly related to C stocks as previously
defined in ForC, with ``\emph{delta.}'' added in front of the variable
name. Further, we added variables capturing the belowground component of
woody mortality (\emph{woody.mortality\_root}) and the combined
aboveground and belowground components of woody mortality
(\emph{woody.mortality}). Although most of these variables lacked
records in ForC as of July 28, 2023, their addition gave the structure
such that records can be populated over time. Finally, to provide better
definition of the previously existing variable \emph{organic.layer},
which has a nebulous definition that reflects the varied definitions
adopted by original studies, we added two clearly defined variables:
\emph{litter} (relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

To submit complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data submission were that (1)
records had been checked against the original study and determined to be
complete and correct, and as originally presented, (2) the original
study presented values in tables or text, as opposed to the values
having been digitized from graphs or calculated based on related
variables, and (3) the records had not previously been submitted to
EFDB. Once converted into EFDB format, the records were reviewed and
then sent to the IPCC's Technical Support Unit for submission to EFDB.
Complete records needed to be reviewed by the EFDB editorial board and
then posted in the database -- a process that lags behind submission of
records and had not yet been completed for all records sent as of July
28, 2023.

@@

As of July 28, 2023, ForC (v4.0) contained 32686 independent records
(39848 total), 19316 of which were for the 42 variables relevant to EFDB
(Fig. 1). These records were distributed across all forested continents
and ecozones, albeit unevenly (Fig. 2). The largest number of records
came from Asia, followed by North America, South America, and Europe,
with relatively few records from Africa, Australia, and Oceania (Fig.
3c). Categorized by FAO ecozone, the greatest numbers of records came
from subtropical humid forests, temperate mountain systems, and tropical
rain forests, each with \textgreater2,000 independent records (Fig. 3b).
Boreal coniferous forests, temperate continental forests, subtropical
mountain systems, and tropical moist deciduous forests had
\textgreater1,000 independent records each, while other ecozones all had
\textless1,000 records. The most widely represented forest type was
needleleaf evergreen, followed by broadleaf deciduous and broadleaf
evergreen (Fig. 3a). In terms of stand age, the most represented age
class was 20-100 years, followed by \textless20 years and then
\textgreater100 years (Fig. 3d).

@@

As of July 28, 2023, we had reviewed or added 2292 EFDB-relevant
records, 1438 records of which were submitted to EFDB, and 376 of which
have been reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
attenuation between records reviewed and those sent to EFDB was
attributable to the presence of digitized records and records where a
variable's value had been calculated as the sum or difference of related
variables rather than presented directly in the text. The discrepancy
between the number of records sent and that posted to EFDB is primarily
attributable to the time required for the IPCC to review and post the
records, and also because a minority of records were deemed not
applicable to EFDB by the review panel.

@@

The ForC database contains large numbers of records that could
potentially be useful for estimating C stock changes applying
methodological guidance provided by the IPCC. Here we have developed a
framework for submitting these records to the EFDB, thus making those
data more accessible for reporting CO\textsubscript{2} emissions and
removals from forest land consistent with good practice in the IPCC
guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of July 28, 2023,
we have submitted 1438 records to EFDB. Although this represents just
7\% of relevant records in ForC, it substantially increases the number
of forest land records in EFDB. The records submitted to EFDB and
present in ForC are very unevenly distributed across variables, regions,
and forest types (Figs. 2-3, Table 2), reflecting broader patterns in
allocation of research effort.

@@

\codedataavailability{All code and data are openly available. The ForC
database and associated code are available via the ForC repository
within the ForC-db organization on GitHub
(https://github.com/forc-db/ForC), and the version used here (ForC v4.0)
is archived in Zenodo \citep[DOI:
10.5281/zenodo.8020861]{anderson-teixeira_forcdb_2023}. The data and
code associated with data submission to EFDB and preparation of this
manuscript are available via the the IPCC-EFDB-integration repository
within the ForC-db organization on GitHub
(https://github.com/forc-db/IPCC-EFDB-integration) and archived in
Zenodo (DOI:
10.5281/zenodo.8021474).} %% use this section when having data sets and software code available
